Fourier-domain modulations and delays of gravitational-wave signals

### 2018-06-18 docs/paper_method.tex
Minor edits in appendicies.

In this Appendix, we summarize our conventions for the Wigner matrices and give a brief description of the construction of a precessing-frame directly from the waveform.

If the $h_{\ell m}^{\rm I}$ are the spin-weighted spherical harmonics~\eqref{eq:defmodes} of the waveform in a fixed inertial frame (I), and if the $h_{\ell m}^{\rm P}$ are the modes of the waveform in a time-dependent precessing frame (P) constructed from the inertial frame by an active rotation with Euler angles $(\alpha, \beta, \gamma)$ (in the convention $(z,y,z)$), then the modes are related by
\begin{subequations}
\label{eq:wignerrotApp}
\begin{align}
	h_{\ell m}^{\rm I} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell *}_{mm'} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm P} \,, \\
	h_{\ell m}^{\rm P} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell }_{m'm} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm I} \,.
\end{align}
\end{subequations}
Here we introduced Wigner $\calD$-matrices
\be\label{eq:defWignerDapp}
	\calD^{\ell}_{mm'} (\alpha, \beta, \gamma) = e^{im \alpha} d^{\ell}_{mm'}(\beta) e^{im' \gamma}\,,
\ee
with the real-valued Wigner $d$-matrix reading
\begin{widetext}
\be\label{eq:defWignerdapp}
	d^{\ell}_{mm'}(\beta) = \sum\limits_{k=k_{\rm min}}^{k_{\rm max}} \frac{(-1)^{k}}{k!} \frac{\sqrt{(l+m)! (l-m)! (l+m')! (l-m')!}}{(l+m-k)! (l-m'-k)! (k-m+m')!} \left( \cos\frac{\beta}{2} \right)^{2\ell+m-m'-2k} \left( \sin\frac{\beta}{2} \right)^{2k-m+m'}\,,
\ee
\end{widetext}
where the boundaries of the sum, $k_{\rm min} = \mathrm{max}(0, m-m')$ and $k_{\rm max} = \mathrm{min}(\ell+m, \ell-m')$, can also be read by enforcing that the arguments of the factorials must be non-negative. Note that our convention differs from the convention of~\cite{ABFO09} by a transposition,
\be
	\calD^{\ell}_{m m'} (\alpha, \beta, \gamma) = D^{\ell \, \mathrm{ABFO}}_{m' m} (\alpha, \beta, \gamma) \,.
\ee
In Sec.~\ref{sec:precession}, we considered only $\ell = 2$ and we restricted ourselves to the contributions of the precessing-frame mode $h_{22}^{\rm P}$. In that case, the relevant explicit expression for the $d$-matrix are
\begin{subequations}
\begin{align}
	d^{2}_{22} (\beta) &= \cos^{4} \frac{\beta}{2}\,, \\
	d^{2}_{12} (\beta) &= 2 \cos^{3} \frac{\beta}{2} \sin \frac{\beta}{2}\,, \\
	d^{2}_{02} (\beta) &= \sqrt{6} \cos^{2} \frac{\beta}{2} \sin^{2} \frac{\beta}{2} \,, \\
	d^{2}_{-12} (\beta) &= 2 \cos \frac{\beta}{2} \sin^{3} \frac{\beta}{2} \,, \\
	d^{2}_{-22} (\beta) &= \sin^{4} \frac{\beta}{2} \,.
\end{align}
\end{subequations}

@@

\section{Explicit expression for the stencil coefficients}
\label{app:stencil}

@@

In this Appendix, we give explicit expressions for the stencil coefficients entering~\eqref{eq:stencilfresnel}. In this work we used only low order stencils with $N\leq 20$, and one can trivially invert of the system~\eqref{eq:stencilsystem}, as this operation has to be done only once. One can also obtain closed-form expressions for these coefficients, thanks to the particular choice of samples at $\pm kT$ that gives to the system~\eqref{eq:stencilsystem} the form of a Vandermonde system. Defining the Vandermonde matrix $V(x_{0},\dots,x_{N})$ as $V_{ij} = (x_{j})^{i}$, setting $x_{j} = j^{2}$, and $b_{p} \equiv (-i\epsilon)^{p}(2p-1)!!$, the linear system~\eqref{eq:stencilsystem} becomes
\be
	b_{p} = \sum\limits_{k=0}^{N} V_{pk} a_{N,k}^{\epsilon} \,.
\ee
%Now, since the Vandermonde system can be seen as a reformulation of the polynomial interpolation problem, we know that $({}^{t}V^{-1})_{ij}$ is the coefficient of $X^{i}$ in the Lagrange interpolation polynomial $\prod_{k\neq j} (X-x_{k})/(x_{j} - x_{k})$. If we introduce the symmetric polynomials $\sigma$ such that
%\be	
%	\sigma(m, \{x_{1}, \dots, x_{n}\}) = \sum\limits_{1\leq i_{1}<\dots<i_{m}\leq n} x_{i_{1}}\dots x_{i_{m}} \,,
%\ee
%we obtain for the inverse of the matrix $V$
%\be
%	V^{-1}_{ij} = \frac{1}{\prod_{k\neq i} (i^{2} - k^{2})} (-1)^{N-j} \sigma(N-j, \{0,\dots,N^{2}\}\backslash \{i^{2}\}) \,.
%\ee
The expression of the inverse of a Vandermonde matrix in terms of symmetric polynomials then allows us to write the $a_{N,k}^{\epsilon}$, for every finite order $N$, as:
\begin{align}
	a_{N,k}^{\epsilon} &= \frac{1}{\prod\limits_{\substack{q=0 \\ q\neq k}}^{N} (q^{2}-k^{2})} \sum\limits_{p=0}^{N} (i\epsilon)^{p}(2p-1)!! \nn\\ & \quad \cdot \sum\limits_{\substack{ 0 \leq j_{1} < \dots < j_{N-p} \leq N \\ j_{1}, \dots, j_{N-p} \neq k}} j_{1}^{2}\dots j_{N-p}^{2}
\end{align}
The two cases $\epsilon\pm 1$ correspond simply to a complex conjugation of the coefficients $a_{N,k}^{\epsilon}$. Table~\ref{tab:stencil} gives the resulting complex rational values for these coefficients for $N\leq 5$.

@@

In this Appendix, we give a more detailed overview of the treatment of precession in existing waveform models that generate signals directly in the Fourier domain. Our objective is not to give an exhaustive account of the various existing models, but rather to highlight how their treatment relates to ours.

@@

In this Appendix, we present an argument for the applicability of the treatment of Section~\ref{sec:formalism} to a Fourier-domain waveform that already contains precession effects, thus combining our investigations of Secs.~\ref{sec:LISA} and~\ref{sec:precession}. Let us first note that one possible approach would be to incorporate both the precession modulation and the LISA modulation and delay in a single $G(f,t)$ kernel. This is what is done notably in~\cite{Klein+15}, with inspiral-only precessing waveforms (and a low-frequency approximation for the response). In that approach, comparing the results of Secs.~\ref{sec:LISA} and Sec.~\ref{sec:precession}, it appears that the precession is more challenging to model than the LISA response -- thus the accuracy of the final result should depend mainly on the precession treatment. In the following, we take a different route and investigate wether we can directly process Fourier-domain precessing mode contributions of the form $\calT^{\ell}_{m m'}(f) \tilde{h}_{\ell m'}(f)$ through the LISA response. We will make simplifying assumptions, assuming simple precession and keeping to order-of-magnitude estimates only. We leave for future work a more thorough investigation of precessing waveforms for LISA beyond simple precession and for IMR waveforms.

>>>>

In this appendix, we summarize our conventions for the Wigner matrices and give a brief description of the construction of a precessing-frame directly from the waveform.

If the $h_{\ell m}^{\rm I}$ are the spin-weighted spherical harmonics~\eqref{eq:defmodes} of the waveform in a fixed inertial frame (I), and if the $h_{\ell m}^{\rm P}$ are the modes of the waveform in a time-dependent precessing frame (P) constructed from the inertial frame by an active rotation with Euler angles $(\alpha, \beta, \gamma)$ (in the convention $(z,y,z)$), then the modes are related by
\begin{subequations}
\label{eq:wignerrotApp}
\begin{align}
	h_{\ell m}^{\rm I} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell *}_{mm'} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm P} \,, \\
	h_{\ell m}^{\rm P} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell }_{m'm} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm I} \,.
\end{align}
\end{subequations}
Here we introduced Wigner $\calD$-matrices
\be\label{eq:defWignerDapp}
	\calD^{\ell}_{mm'} (\alpha, \beta, \gamma) = e^{im \alpha} d^{\ell}_{mm'}(\beta) e^{im' \gamma}\,,
\ee
with the real-valued Wigner $d$-matrix reading
\begin{widetext}
\be\label{eq:defWignerdapp}
	d^{\ell}_{mm'}(\beta) = \sum\limits_{k=k_{\rm min}}^{k_{\rm max}} \frac{(-1)^{k}}{k!} \frac{\sqrt{(l+m)! (l-m)! (l+m')! (l-m')!}}{(l+m-k)! (l-m'-k)! (k-m+m')!} \left( \cos\frac{\beta}{2} \right)^{2\ell+m-m'-2k} \left( \sin\frac{\beta}{2} \right)^{2k-m+m'}\,,
\ee
\end{widetext}
where the boundaries of the sum, $k_{\rm min} = \mathrm{max}(0, m-m')$ and $k_{\rm max} = \mathrm{min}(\ell+m, \ell-m')$, can also be read by enforcing that the arguments of the factorials must be non-negative. Note that our convention differs from the convention of~\cite{ABFO09} by a transposition,
\be
	\calD^{\ell}_{m m'} (\alpha, \beta, \gamma) = D^{\ell \, \mathrm{ABFO}}_{m' m} (\alpha, \beta, \gamma) \,.
\ee
In Sec.~\ref{sec:precession}, we considered only $\ell = 2$ and we restricted ourselves to the contributions of the precessing-frame mode $h_{22}^{\rm P}$. In that case, the relevant explicit \jgb{expressions for the $d$-matrix elements} are
\begin{subequations}
\begin{align}
	d^{2}_{22} (\beta) &= \cos^{4} \frac{\beta}{2}\,, \\
	d^{2}_{12} (\beta) &= 2 \cos^{3} \frac{\beta}{2} \sin \frac{\beta}{2}\,, \\
	d^{2}_{02} (\beta) &= \sqrt{6} \cos^{2} \frac{\beta}{2} \sin^{2} \frac{\beta}{2} \,, \\
	d^{2}_{-12} (\beta) &= 2 \cos \frac{\beta}{2} \sin^{3} \frac{\beta}{2} \,, \\
	d^{2}_{-22} (\beta) &= \sin^{4} \frac{\beta}{2} \,.
\end{align}
\end{subequations}

@@

\section{Explicit expression for the \jgb{Fresnel} stencil coefficients}
\label{app:stencil}

@@

In this appendix, we give explicit expressions for the stencil coefficients entering~\eqref{eq:stencilfresnel} \jgb{which approximates a Fresnel transform, as discussed in Sec.~\ref{subsec:fresneltransform}}. In this work we used only low order stencils with $N\leq 20$, and one can trivially invert of the system~\eqref{eq:stencilsystem}, as this operation has to be done only once. One can also obtain closed-form expressions for these coefficients, thanks to the particular choice of samples at $\pm kT$ that gives to the system~\eqref{eq:stencilsystem} the form of a Vandermonde system. Defining the Vandermonde matrix $V(x_{0},\dots,x_{N})$ as $V_{ij} = (x_{j})^{i}$, setting $x_{j} = j^{2}$, and $b_{p} \equiv (-i\epsilon)^{p}(2p-1)!!$, the linear system~\eqref{eq:stencilsystem} becomes
\be
	b_{p} = \sum\limits_{k=0}^{N} V_{pk} a_{N,k}^{\epsilon} \,.
\ee
%Now, since the Vandermonde system can be seen as a reformulation of the polynomial interpolation problem, we know that $({}^{t}V^{-1})_{ij}$ is the coefficient of $X^{i}$ in the Lagrange interpolation polynomial $\prod_{k\neq j} (X-x_{k})/(x_{j} - x_{k})$. If we introduce the symmetric polynomials $\sigma$ such that
%\be	
%	\sigma(m, \{x_{1}, \dots, x_{n}\}) = \sum\limits_{1\leq i_{1}<\dots<i_{m}\leq n} x_{i_{1}}\dots x_{i_{m}} \,,
%\ee
%we obtain for the inverse of the matrix $V$
%\be
%	V^{-1}_{ij} = \frac{1}{\prod_{k\neq i} (i^{2} - k^{2})} (-1)^{N-j} \sigma(N-j, \{0,\dots,N^{2}\}\backslash \{i^{2}\}) \,.
%\ee
The expression of the inverse of a Vandermonde matrix in terms of symmetric polynomials then allows us to write the $a_{N,k}^{\epsilon}$, for every finite order $N$, as:
\begin{align}
	a_{N,k}^{\epsilon} &= \frac{1}{\prod\limits_{\substack{q=0 \\ q\neq k}}^{N} (q^{2}-k^{2})} \sum\limits_{p=0}^{N} (i\epsilon)^{p}(2p-1)!! \nn\\ & \quad \cdot \sum\limits_{\substack{ 0 \leq j_{1} < \dots < j_{N-p} \leq N \\ j_{1}, \dots, j_{N-p} \neq k}} j_{1}^{2}\dots j_{N-p}^{2}
\end{align}
The two cases $\epsilon\pm 1$ correspond simply to a complex conjugation of the coefficients $a_{N,k}^{\epsilon}$. Table~\ref{tab:stencil} gives the resulting complex rational values for these coefficients for $N\leq 5$.

@@

In this appendix, we give a more detailed overview of the treatment of precession in existing waveform models that generate signals directly in the Fourier domain. Our objective is not to give an exhaustive account of the various existing models, but rather to highlight how their treatment relates to ours.

@@

In this appendix, we present an argument for the applicability of the treatment of Section~\ref{sec:formalism} to a Fourier-domain waveform that already contains precession effects, thus combining our investigations of Secs.~\ref{sec:LISA} and~\ref{sec:precession}. Let us first note that one possible approach would be to incorporate both the precession modulation and the LISA modulation and delay in a single $G(f,t)$ kernel. This is what is done notably in~\cite{Klein+15}, with inspiral-only precessing waveforms (and a low-frequency approximation for the response). In that approach, comparing the results of Secs.~\ref{sec:LISA} and Sec.~\ref{sec:precession}, it appears that the precession is more challenging to model than the LISA response -- thus the accuracy of the final result should depend mainly on the precession treatment. In the following, we take a different route and investigate wether we can directly process Fourier-domain precessing mode contributions of the form $\calT^{\ell}_{m m'}(f) \tilde{h}_{\ell m'}(f)$ through the LISA response. We will make simplifying assumptions, assuming simple precession and keeping to order-of-magnitude estimates only. We leave for future work a more thorough investigation of precessing waveforms for LISA beyond simple precession and for IMR waveforms.
